# Intro to data wrangling, Part II

__WARNING: Content subject to change__

```{r}
suppressPackageStartupMessages(library(tidyverse))
```


__Announcements__:

- Peer review 1 due tonight.
- Homework 2 due tonight.
- stat545.com
- Reminder about setting up your own hw repo.

__Follow-up__: 

- By the way, STAT 545 jumps into the tidyverse way of doing things in R, instead of the base R way of doing things. Lecture 2 was about "just enough" base R to get you started.
- `if` statement on its own won't work within `mutate()` because it's not vectorized. Would need to vectorize it with `sapply()` or, better, the `purrr::map` family:

```{r}
tibble(a = 1:4) %>% 
  mutate(b = if (a < 3) "small" else "big",
         c = sapply(a, function(x) if (x < 3) "small" else "big"),
         d = map_chr(a, ~ if(.x < 3) "small" else "big"))
```



## Today's Lessons

Today's topic is _one-table verbs with `dplyr`_. We've already seen `select`, `filter`, and `arrange`. 

- mutate
- grouped summarize
- grouped mutate

...and their corresponding theoretical function types. 

tsibbles

## Resources

Concepts from today's class are closely mirrored by the following resources.

- Jenny's tutorial on [all the one-table verbs](http://stat545.com/block010_dplyr-end-single-table.html)

Other resources:

- Like learning from a textbook? Check out all of [r4ds: transform](http://r4ds.had.co.nz/transform.html).
- The [intro to `dplyr` vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) is also a great resource. 

Resources for specific concepts:

- To learn more about window functions and how dplyr handles them, see the [window-functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html) vignette for the `dplyr` package. 

## Participation

To get participation points for today, we'll be filling out the cm008-exercise.Rmd file ([Rmd](https://github.com/STAT545-UBC/Classroom/blob/master/notes/cm008-exercise.Rmd), or [from html](http://stat545.com/Classroom/notes/cm008-exercise.nb.html)), and adding it to your participation repo. 

Let's take time to get set up:

1. Download the file as `.Rmd`. 
2. Move the file to your local participation repo.
3. Open the `.Rmd` file, and change the output type to `github_document`.
4. _Optional, but recommended_: Stage and commit the Rmd file (you can do this through RStudio).
    - You can still get participation marks by uploading the final files to GitHub.
