# Intro to plotting with `ggplot2`, Part II

__WARNING: Content subject to change__

__Today__: Continue with `ggplot2`.
    - This time, exploration of aesthetics (and facetting). 
    - Continue with geom exploration if time permits. 


Note that ggplot2's grouping is different from dplyr's! Try making a spaghetti plot of lifeExp over time for each coutry, by piping in a grouped data frame -- it won't work:

```{r}

```

## Plot types / `geom`s

Here, we'll explore some common plot types, and how to produce them with `ggplot2`.

### Histograms: `geom_histogram()`

Useful for depicting the distribution of a continuous random variable. Partitions the number line into bins of certain width, counts the number of observations falling into each bin, and erects a bar of that height for each bin.

Required aesthetics:

- `x`: A numeric vector.

By default, a histogram plots the _count_ on the y-axis. If you want to use proportion, specify the `y = ..prop..` aesthetic. 

You can change the smoothness of the plot via two arguments (your choice):

- `bins`: the number of bins/bars shown in the plot.
- `binwidth`: the with of the bins shown on the plot.

Example:

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(bins = 50)
```


### Density: `geom_density()`

Essentially, a "smooth" version of a histogram. Uses [kernels](https://en.wikipedia.org/wiki/Kernel_density_estimation) to produce the curve.

Required aesthetics:

- `x`: A numeric vector.

Good to know:

- `bw` argument controls the smoothness: Smaller = rougher.

Example:

```{r}
ggplot(gapminder, aes(lifeExp)) +
  geom_density()
```

### Jitter plots: `geom_jitter()`

A scatterplot, but with minor random perturbations of each point. Useful for scatterplots where points are overlaying, or when one variable is categorical.

Required aesthetics:

- `x`: any vector
- `y`: any vector

Example:

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_jitter()
```

### Box plots: `geom_boxplot()`

This geom makes a boxplot for a numeric variable in each of a category. Useful for visualizing probability distributions across different categories.

Required aesthetics:

- `x`: A factor (categorical variable)
- `y`: A numeric variable

Example:

```{r}
ggplot(gapminder, aes(continent, lifeExp)) +
  geom_boxplot()
```


### Ridge plots: `ggridges::geom_density_ridges()`

A (superior?) alternative to the boxplot, the ridge plot (also known as the joy plot) places a kernel density for each group, instead of the box. 

You'll need to install the `ggridges` package. You can do lots more with ridges -- check out [the ggridges intro vignette](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html).

Required aesthetics (reversed from boxplots!)

- `x`: A numeric variable
- `y`: A factor (categorical variable) 

Example:

```{r}
ggplot(gapminder, aes(lifeExp, continent)) +
  ggridges::geom_density_ridges(bins = 50)
```

### Bar plots: `geom_bar()` or `geom_col()`

These geom's erect a bar over each category.

`geom_bar()` automatically determines the height of the bar according to the count of each category.

`geom_col()` requires a manual specification of the bar heights.

Required aesthetics:

- `x`: A categorical variable
- `y`: A numeric variable (only required for `geom_col()`!)
  - To use proportion in `geom_bar()` instead of count, set `y = ..prop..`

Example: number of 4-, 6-, and 8- cylinder cars in the `mtcars` dataset:

```{r}
ggplot(mtcars, aes(cyl)) +
  geom_bar()
```

### Line charts: `geom_line()`

A line plot connects points with straight lines, from left-to-right. Especially useful if time is on the x-axis.

Required aesthetics:

- `x`: a variable having some ordering to it.
- `y`: a numeric variable.

Although not required, the `group` aesthetic will come in handy here. This aesthetic produces a plot independently for each group, and overlays the results.


### Path plots: `geom_path()`

Like `geom_line()`, except connects points in the order that they appear in the dataset.

### Activity: Fix the Plots

Fill out __Exercise 3: Fix the plots (Together)__ in your worksheet.

_Bus stop_: Did you lose track of where we are? You can still do the exercise!

1. Click [here](https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/tutorials/cm005-exercise.Rmd) to obtain the worksheet if you don't have it.
2. Set up the workspace by following the instructions in the "Preliminary" section.
3. You're all set! Hint for completing the exercise: use the information from this section ("Plot types / `geom`s") to complete the exercise.

