# (4) Functional programming in R: Part II

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, warning = FALSE, error = FALSE}
library(tidyverse)
library(rlang)
library(glue)
#install.packages('docopt')
#install.packages('testthat')
library(docopt) # you will need to install this using install.packages('docopt')
library(testthat) # you will need to install this using install.packages('testthat')
```

## Today's Agenda

- Announcements:
  - If you need Git help, please come to our [office hours](https://stat545.stat.ubc.ca/officehours/)! This is a crucial concept and it's worth getting it right!
  - Your project repos should be *public*, assignment and participation repos are *private*
    - Apologies if I mis-spoke and for any confusion. See [GitHub issue](https://github.com/STAT547-UBC-2019-20/Discussions/issues/17) for more
  - Next Thursday's class

- Part 1: Running Rscripts from the terminal/command prompt (20 mins)
  - "REPL" vs. non-interactive scripts
  - Converting interactive R code to a scripted, non-interactive version
  
- Part 2: Command-line arguments and docopt (30 mins)
  - Motivating command-line arguments
  - Options for specifying command-line arguments
  - Adding command-line arguments to your R scripts

- Part 3: Writing tests for your functions (25 mins)
  - Motivating the use of tests in programming
  - Introducing the `testthat` package
  - Writing tests using the `testthat` package

## Learning outcomes for this lecture

1. Evaluate when it is optimal to code in a REPL framework (interactively) vs. using scripts. 
1. Demonstrate the ability to run interactive R code from a .R script.
1. Use principles of functional programming to create functions inside R scripts.
1. Explain the purpose of command-line arguments and describe the various flavours (optional, required, long and short forms, repeating, etc...).
1. Use the `docopt` to create commandline arguments for R scripts.
1. Defend the use of tests when programming.
1. Use the `testthat` library to write tests for R functions.

## Part 1: Running Rscripts from the terminal/command prompt (15 mins)

This section has been adapted from Tiffany Timbers' DSCI 522 lectures found [here](https://github.com/UBC-MDS/DSCI_522_dsci-workflows/blob/master/lectures/02_lecture-intro-to-scripts.ipynb).

### "REPL" vs. non-interactive scripts
So far in STAT 545 and 547M, we have only been working on Rmd files and interactively working with code.
We run and re-run specific code chunks, add bits of code, and re-run it to get updated results. 
Of course, this works - but is it reproducible ?
Can you trust that in a week, or a month, or a year from now that you will be able to reproduce the exact steps needed to get the same answer?
Usually not. 
The above style of working actually has a name: REPL, or Read-Evaluate-Print-Loop.
Despite the fact that it's usually not easily reproducible, it is actually VERY useful!
It's good for:
  - solving smaller problems in projects limited in scope
  - explore and play around with unfamiliar and unknown datasets
  - developing code that will eventually make its way into a script or turned into a function
  - developing code that you will only use once
It is rare that data scientists would go directly towards creating scripts and functions without first developing them in a REPL environment. 

At this point, you may be wondering what scripts *are* and *why* we should write them?.
Well, an R script is just a text file that contains R code that is run in a sequence, usually with no or minimal intervention from the user. 
Scripts are generally run from the Terminal, but can also be run from IDEs (interactive developer environment; RStudio is an IDE) and even the Rconsole using `source(script_name.R)`.

As for why:
  - Scripts are generally more time efficient (especially for long term usage)
  - You can combine multiple scripts to create an analysis pipeline (next week!)
  - Scripts can be reused and adapted for different purposes; additional functionality can be added to scripts without removing the ability of it working for previous analyses
  - Are an important component of reproducible workflows

Alright now that we know how REPL, or interactive scripts differ from non-interactive scripts, let's move on and actually create R scripts from interactive R code.

### Converting interactive R code to a scripted, non-interactive version

Let's just do a quick sanity check to make sure our script can run. 
- Step 1: In the cm104 directory, create a new blank .R script and save it as `first_script.R`
  - In RStudio, File > New File > R Script.
- Step 2: Write the following R code inside `first_script.R`:
  - `print('Hello World')`
- Step 3: Run the Rscript by opening a new Terminal, navigating to the directory where `first_script.R` exists and then type:
  - `Rscript first_script.R`
  - "Hello World" should be printed to the Terminal and the R script will finish.
- Step 4: Navigate to one directory above the current directory with `cd ..`
  - Now try running `first_script.R` - does it work?
  - How would you modify `Rscript first_script.R` to make it work (without going into the `cm104` directory)?

```
## YOUR SOLUTION HERE
Rscript cm104/first_script.R
```
  
Okay, let's do something slightly more interesting: a quick demo using an interactive script that we're going to write together.

#### Your Task 1: 
  - Start with the `mtcars` dataset
  - Use `dplyr::mutate` to create a new column for the fuel efficiency in L/100km rather than mpg
  - Create a plot of Horse power vs. fuel efficiency

```{r interactive_mtcars_demo}
head(mtcars)
## YOUR SOLUTION HERE
mtcars %>%
  mutate(lkm = 235.2145 / mpg) %>% # imperial = 282.48, #US = 235.21
  ggplot(aes(x = lkm, y = hp)) + geom_point() + 
  theme_bw(16) + # Set font size like this 
  labs(x = 'Fuel efficiency (L/100 km)',
       y = 'Horse power (hp)',
       title = "Fuel efficiency by horse power")
```

#### Your Task 2: 
  - Use the code above as a starting point
  - Copy the code above and paste it into `first_script.R`
  - Move the mutate command into a separate function
  - Your function should take in a string argument with two options: "USA" and "Imperial" (set a reasonable default)
  - Try running the script. Did it work? Was there a plot produced?
  - If yes, great! If not, what did you have to change to get it to work? 
  - Can you explain why?
  
```{r working_mtcars_demo}
## YOUR SOLUTION HERE
library(tidyr)
library(dplyr)
library(ggplot2)

fuel_conversion <- function (df, gallon_type = 'USA') { 
  
  if (gallon_type == 'USA') {
    conv = 235.2145
  } else {
    conv = 282.48
  }
  df %>% mutate(lkm = conv / mpg)
    
  }

mtcars %>%
  fuel_conversion(gallon_type = 'Imperial') %>%
  ggplot(aes(x = lkm, y = hp)) + geom_point() +
  theme_bw(16) + # Set font size like this
  labs(x = 'Fuel efficiency (L/100 km)',
       y = 'Horse power (hp)',
       title = "Fuel efficiency by horse power") +
  ggsave('test_plot.png', width = 8, height = 5)

```

Great! Looks like things are working!
What if we wanted to specify the arguments of assignment to the imperial definition of gallon rather than the US one?
For that, we need to look at command line arguments!

## Part 2: Command-line arguments and docopt (30 mins)



## Part 3: Writing tests for your functions

This section has been adapted from 

This section is adapted from Chapter 8 of the [tidynomicon textbook by Greg Wilson](https://gvwilson.github.io/tidynomicon/).

### Testing and Error Handling

Novices write code and pray that it works.
Experienced programmers know that prayer alone is not enough, and take steps to protect what little sanity they have left.
This chapter looks at the tools R gives us for doing this.

### How does R handle errors?

We say that the operation [signals](glossary.html#signal-condition) a [condition](glossary.html#condition) that some other piece of code then [handles](glossary.html#handle-condition).
These things are all simpler to do using the rlang library, so we begin by loading that:

In order of increasing severity, the three built-in kinds of conditions are [messages](glossary.html#message), [warnings](glossary.html#warning), and [errors](glossary.html#error).
(There are also interrupts, which are generated by the user pressing Ctrl-C to stop an operation, but we will ignore those for the sake of brevity.)
We can signal conditions of these kinds using the functions `message`, `warning`, and `stop`, each of which takes an error message as a parameter:

```{r message-warning-error, error=TRUE}
message("This is a message.")
warning("This is a warning.\n")
stop("This is an error.")
```

Note that we have to supply our own line ending for warnings but not for the other two cases.
Note also that there are very few situations in which a warning is appropriate: if something has truly gone wrong then we should stop, but otherwise we should not distract users from more pressing concerns.

The bluntest of instruments for handling errors is to ignore them. 
If a statement is wrapped in the function `try` then errors that occur in it are still reported, but execution continues.
Compare this:

```{r attempt-without-try, error=TRUE}
attemptWithoutTry <- function(left, right){
  temp <- left + right
  "result" # returned
}
result <- attemptWithoutTry(1, "two")
cat("result is", result)
```

with this:

```{r attempt-using-try}
attemptUsingTry <- function(left, right){
  temp <- try(left + right)
  "value returned" # returned
}
result <- attemptUsingTry(1, "two")
cat("result is", result)
```

We can suppress error messages from `try` by setting `silent` to `TRUE`:

```{r attempt-quietly}
attemptUsingTryQuietly <- function(left, right){
  temp <- try(left + right, silent = TRUE)
  "result" # returned
}
result <- attemptUsingTryQuietly(1, "two")
cat("result is", result)
```

Do NOT do this, lest you one day find yourself lost in a silent hellscape.

Should you more sensibly wish to handle conditions rather than ignore them, you may invoke `tryCatch`.

We begin by raising an error explicitly:

```{r r-try-catch}
tryCatch(
  stop("our message"),
  error = function(cnd) print(glue("error object is {cnd}"))
)
```
We can now run a function that would otherwise blow up:

```{r r-try-catch-triggered}
tryCatch(
  attemptWithoutTry(1, "two"),
  error = function(cnd) print(glue("error object is {cnd}"))
)
```

### What should I know about testing in general?

In keeping with common programming practice, we have left testing until the last possible moment.
The standard testing library for R is [testthat][testthat] and this is how tests are handled in R:

1.  Each test consists of a single function that tests a single property or behavior of the system.
2.  Tests are collected into files with prescribed names that can be found by a test runner (code that is automatically run once after each unit test).
3.  Shared setup (code that is automatically run once before each unit test) and teardown (Code that is automatically run once after each unit test) steps are put in functions of their own.

For now in STAT547, we will use the `testthat` library to test our functions in-line our Rscripts, though it is much better practice to set up your tests in a `tests` directory and have them run automatically.

Let's load it and write our first test:

```{r introduce-testthat}
test_that("Zero equals itself", {expect_equal(0, 0)})
```

As is conventional with unit testing libraries, no news is good news: if a test passes, it doesn't produce output because it doesn't need our attention.
Let's try something that ought to fail:

```{r error = TRUE}
test_that("Zero equals one", {expect_equal(0, 1)})
```

If you run this, you should see something like:

```
Error: Test failed: 'Zero equals one' * 0 not equal to 1. 1/1 mismatches [1] 0 - 1 == -1
```

Good: we can draw some comfort from the fact that Those Beyond have not yet changed the fundamental rules of arithmetic.
But what are the curly braces around `expect_equal` for? The answer is that they create a code block for `test_that` to run.
We can run `expect_equal` on its own:

```{r expect-equal-alone, error=TRUE}
expect_equal(0, 1)
```

but that doesn't produce a summary of how many tests passed or failed.
Passing a block of code to `test_that` also allows us to check several things in one test:

```{r pass-code-block, error=TRUE}
test_that("Testing two things", {
  expect_equal(0, 0)
  expect_equal(0, 1)
})
```

A block of code is *not* the same thing as an anonymous function, which is why running this block of code does nothing — the "test" defines a function
but doesn't actually call it:

```{r anonymous-function}
test_that("Using an anonymous function", function() {
  print("In our anonymous function")
  expect_equal(0, 1)
})
```

### How should I organize my tests?

Running blocks of tests by hand is a bad practice.
Instead, we should put related tests in files and then put those files in a directory called `tests/testthat`.
We can then run some or all of those tests with a single command.

To start, create in your `cm104` folder, create this file: `./tests/testthat/test_example.R`:

```
library(testthat)
context("Demonstrating the testing library")

test_that("Testing a number with itself", {
  expect_equal(0, 0)
  expect_equal(-1, -1)
  expect_equal(Inf, Inf)
})

test_that("Testing different numbers", {
  expect_equal(0, 1)
})

test_that("Testing with a tolerance", {
  expect_equal(0, 0.01, tolerance = 0.05, scale = 1)
  expect_equal(0, 0.01, tolerance = 0.005, scale = 1)
})
```

The first line loads the testthat package, which gives us our tools.
The call to `context` on the second line gives this set of tests a name for reporting purposes.
After that, we add as many calls to `test_that` as we want, each with a name and a block of code.
We can now run this file from within RStudio:

```
test_dir("tests/testthat")
```

The output should be: 

```
✔ |  OK F W S | Context

⠏ |   0       | Skipping rows correctly
✖ |   0 5     | Skipping rows correctly
───────────────────────────────────────────────────────────────────────────
test_determine_skip_rows_a.R:9: failure: The right row is found when there are header rows
`result` not equal to 2.
Lengths differ: 0 is not 1

test_determine_skip_rows_a.R:14: failure: The right row is found when there are header rows and blank lines
`result` not equal to 3.
Lengths differ: 0 is not 1

test_determine_skip_rows_a.R:19: failure: The right row is found when there are no header rows to discard
`result` not equal to 0.
Lengths differ: 0 is not 1

test_determine_skip_rows_a.R:23: failure: No row is found when 'iso3' isn't present
`determine_skip_rows("a1,a2\nb1,b1\n")` did not throw an error.

test_determine_skip_rows_a.R:28: failure: No row is found when 'iso3' is in the wrong place
`determine_skip_rows("stuff,iso3\n")` did not throw an error.
───────────────────────────────────────────────────────────────────────────

⠏ |   0       | Skipping rows correctly
✔ |   5       | Skipping rows correctly

⠏ |   0       | Demonstrating the testing library
✖ |   4 2     | Demonstrating the testing library
───────────────────────────────────────────────────────────────────────────
test_example.R:11: failure: Testing different numbers
0 not equal to 1.
1/1 mismatches
[1] 0 - 1 == -1

test_example.R:16: failure: Testing with a tolerance
0 not equal to 0.01.
1/1 mismatches
[1] 0 - 0.01 == -0.01
───────────────────────────────────────────────────────────────────────────

⠏ |   0       | Finding empty rows
✖ |   1 2     | Finding empty rows
───────────────────────────────────────────────────────────────────────────
test_find_empty_a.R:9: failure: A single non-empty row is not mistakenly detected
`result` not equal to NULL.
Types not compatible: integer is not NULL

test_find_empty_a.R:14: failure: Half-empty rows are not mistakenly detected
`result` not equal to NULL.
Types not compatible: integer is not NULL
───────────────────────────────────────────────────────────────────────────

⠏ |   0       | Finding empty rows
✔ |   3       | Finding empty rows

⠏ |   0       | Testing properties of tibbles
✔ |   1   1   | Testing properties of tibbles
───────────────────────────────────────────────────────────────────────────
test_tibble.R:6: warning: Tibble columns are given the name 'value'
`as.tibble()` is deprecated, use `as_tibble()` (but mind the new semantics).
This warning is displayed once per session.
───────────────────────────────────────────────────────────────────────────

══ Results ════════════════════════════════════════════════════════════════
Duration: 0.2 s

OK:       14
Failed:   9
Warnings: 1
Skipped:  0
```

Care is needed when interpreting these results.
There are four `test_that` calls, but eight actual checks, and the number of successes and failures is counted by recording the latter, not the former.

What then is the purpose of `test_that`?
Why not just use `expect_equal` and its kin, such as `expect_true`, `expect_false`, `expect_length`, and so on?
The answer is that it allows us to do one operation and then check several things afterward.

Our tests still aren't checking anything statistical, but without trustworthy data,
our statistics will be meaningless.
Tests like these allow our future selves to focus on making new mistakes instead of repeating old ones.

### What type of tests can I do with `testthat` ?

[This page](https://testthat.r-lib.org/reference/index.html) has a thorough list of "expectations".
Some useful functions are:

- test_dir(), test_package(), test_check(), is_testing(), testing_package(): Expectation: is returned value less or greater than specified value?
- expect_equal(), expect_equivalent(), expect_identical(), expect_reference(): Expectation: is the object equal to a value?
- expect_length(): Expectation: does a vector have the specified length?
... etc

### More practice with `test_that`

The [next two sub-sections](http://tidynomicon.tech/testerror.html#how-should-i-organize-my-tests) of the tidynomicon chapter on testing are likely beyond the scope of STAT547, but I highly encourage you to go through on your own time when you're about to embark on a project for your research!

### Key Points of testing

- The three built-in levels of conditions are messages, warnings, and errors.
- Programs/scripts can signal these themselves using the functions `message`, `warning`, and `stop`.
- Operations can be placed in a call to the function tryCatch to handle errors.
- Use `testthat` to write unit tests for R.
- Put tests in files called test_group.R and call them test_something.
- Write tests for data transformation steps (if you can) as well as library functions.

## Additional Resources

- Stat545.com by Jenny Bryan - [Tests](https://stat545.com/functions-part3.html)
- Wickham and Bryan's R Packages book - [Testing chapter](https://r-pkgs.org/tests.html)
- Tidynomicon by Greg Wilson - [Tests in R]