# STAT 547M {-}

STAT 547 is under active development. The materials will be updated evry week prior to the class.

# (1) Review of STAT 545

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

## Today's Agenda

- Part 1: Introductions and course overview (15 mins)
  - Two parts to the course: reproducible workflows and dashboards
  - Course grading elements
  - Group projects
  - Optional "lab"
  - Office hours
  - What to expect in STAT 547
  
- Part 2: Review of STAT545 (45 mins)
  - tidy data
  - factors in R
  - ggplot
  - dplyr
  - File I/O

- Part 3: Group projects in STAT 547 (10 mins)
  - Partners
  - Choosing a research question
  - What's due this week
  
```{r test1}
var <- c(1,2,3,5)
```
  
## Part 1: Introductions and course overview (15 mins)

### Two parts to the course: reproducible workflows and dashboards

STAT 547 can roughly be divided into two main ideas or themes: reproducible workflows and interactive dashboards.
In the first part of the course we will start working with writing functions in R scripts and think a little bit about functional programming.
Then with some new tools at our disposal, we will create "analysis pipelines" to automate and streamline the process going from loading raw data, all the way to creating a final report with the results of a regression analysis.
Next we'll switch gears and use our functional programming skills to create a dashboard app with multiple interactive components and multi-page/tab layouts.
Finally we will end with deploying our dashboard so that it is publicly available!


### Course grading elements
### Group projects
### Optional "lab"
### Office hours
### What to expect in STAT 547

## Part 2: Review of STAT545 (45 mins)

### tidy data
### factors in R
### ggplot
### dplyr
### File I/O

## Part 3: Group projects in STAT 547

### Partners
### Choosing a research question
### What's due this week
  