# (6) Parameterized reports and presentations

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

## Today's Agenda

- Announcements:
  - Firas is away today, Yulia will cover my office hours after lecture in SWNG407
  - Hayley Boyce is covering lecture 6 on Parameterized reports
  - Next week we will switch gears to creating dashboards!

- Part 1: Short demos of RMarkdown documents (20 mins)
  - RMarkdown presentations (with citations and references)
  - RMarkdown templated reports
  - RMarkdown analyses
  
- Part 2: Output options in RMarkdown (20 mins)
  - Review of YAML headers
  - Table of contents & navigating between headings and images
  - Output options (html, md, pdf, word, github )
  
- Break

- Part 3: Parameterized Reporting (30 mins)
  - Global options
  - Code chunk options (`include`, `echo`, `message`, `warning`)
  - How to use R variables in-line with text

- Part 4: Citations in RMarkdown (Optional, if there's time; 10 mins)
  - Bibtex and RMarkdown
  - Citing scientific papers
  - Citing packages

## Part 1: Short demos of RMarkdown documents 

## Part 2: YAML Headers and code chunks (15 mins)

## Part 3: Parameterized reporting (20 mins)

## Part 4: References and citations in RMarkdown (15 mins)