# (12) Introduction to creating packages in R

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
```

**This lecture is 60% complete**

## Today's Agenda

- Announcements: (2 mins)
  - Reminder! Participation repos are due soon!
  - [Assignment 6](https://stat545.stat.ubc.ca/evaluation/assignment_06/assignment_06/) is posted (I have added 24 hours extra since it was posted late)

- Part 1: Showcase of the best STAT 547 dashboards! (10 mins)
  - [Dashboard 1 - by ]()
  - [Dashboard 2 - by ]()
  - [Dashboard 3 - by ]()
  
- Part 2: Package in R (10 mins)
  - Why make packages?
  - Components of a package in R
  - Releasing a package
  
- Part 3: Creating a package in R (45 mins)
  - [Chapter 2 of R packages](https://r-pkgs.org/whole-game.html)
  
- Part 4: Final debrief: what's next? (5 mins)
  - RShiny
  - Tests! Unit Tests!
  - Continuous integration

## Part 1: Showcase of the best STAT 547 dashboards! (10 mins)

## Part 2: Creating a package in R (45 mins)

## Part 3: Creating a package in R (45 mins)

## Part 4: Final debrief: what's next? (5 mins)