---
title: "cm009 Exercises: tidy data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
```

## Reading and Writing Data: Exercises

Make a tibble of letters, their order in the alphabet, and then a pasting of the two columns together.

```{r}

```

Make a tibble of three names and commute times.

```{r}

```


Write the `iris` data frame as a `csv`. 

```{r}

```

Write the `iris` data frame to a file delimited by a dollar sign. 

```{r}

```

Read the dollar-delimited `iris` data to a tibble.

```{r}

```

Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://github.com/jennybc/lotr-tidy/blob/master/data/The_Return_Of_The_King.csv

```{r}

```

## `gather()`

(Exercises largely based off of Jenny Bryan's [gather tutorial](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md))

This function is useful for making untidy data tidy (so that computers can more easily crunch the numbers).

1. Combine the three LOTR untidy tables (`lotr1`, `lotr2`, `lotr3`) to a single untidy table by stacking them.  

```{r}

```

2. Convert to tidy. Also try this by specifying columns as a range, and with the `contains()` function.

```{r}

```

3. Try again (bind and tidy the three untidy data frames), but without knowing how many tables there are originally. 
    - The additional work here does not require any additional tools from the tidyverse, but instead uses a `do.call` from base R -- a useful tool in data analysis when the number of "items" is variable/unknown, or quite large. 

```{r}

```

## `spread()`

(Exercises largely based off of Jenny Bryan's [spread tutorial](https://github.com/jennybc/lotr-tidy/blob/master/03-spread.md))

This function is useful for making tidy data untidy (to be more pleasing to the eye).

Read in the tidy LOTR data (despite having just made it):

```{r}
lotr_tidy <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/lotr_tidy.csv")
```

Get word counts across "Race". Then try "Gender". 

```{r}

```

Now try combining race and gender. Use `unite()` from `tidyr` instead of `paste()`. 

```{r}

```

## Other `tidyr` goodies

Check out the Examples in the documentation to explore the following.

`expand` vs `complete` (trim vs keep everything). Together with `nesting`. Check out the Examples in the `expand` documentation.

```{r}

```

`separate_rows`: useful when you have a variable number of entries in a "cell".

```{r}

```

`unite` and `separate`.

```{r}

```

`uncount` (as the opposite of `dplyr::count()`)

```{r}

```


`drop_na` and `replace_na`

```{r}

```

`fill`

```{r}

```


`full_seq`

```{r}

```


## Time remaining?

Time permitting, do [this exercise](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md#exercises) to practice tidying data. 