---
title: "cm011 Exercises: R as a programming language"
output: html_notebook
---


## Reading and Writing Data: Exercises

Make a tibble of letters, their order in the alphabet, and then a pasting of the two columns together.

```{r}

```

Make a tibble of three names and commute times.

```{r}

```


Write the `iris` data frame as a `csv`. 

```{r}

```

Write the `iris` data frame to a file delimited by a dollar sign. 

```{r}

```

Read the dollar-delimited `iris` data to a tibble.

```{r}

```

Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://github.com/jennybc/lotr-tidy/blob/master/data/The_Return_Of_The_King.csv

```{r}

```


# Part I

## Types and Modes and Classes, Oh My!

R objects have a _type_, a _mode_, and a _class_. This can be confusing:

```{r}
a <- 3
print(typeof(a))
print(mode(a))
print(class(a))
```


```{r}
print(typeof(iris))
print(mode(iris))
print(class(iris))
```


```{r}
print(typeof(sum))
print(mode(sum))
print(class(sum))
```

Usually, there's no need to fuss about these differences: just use the `is.*()` family of functions. Give it a try:

```{r}

```

We can also coerce objects to take on a different form, typically using the `as.*()` family of functions. We can't always coerce! You'll get a sense of this over time, but try:

- Coercing a number to a character.
- Coercing a character to a number.
- Coercing a number to a data.frame. `letters` to a data.frame.

```{r}

```

There is also a slight difference between coercion and conversion, but this is usually not important.

## Vectors

Vectors store multiple entries of a data type. You'll discover that they show up just about everywhere in R, so they're fundamental and extremely important.

### Vector Construction and Basic Subsetting

We've seen vectors as columns of data frames:

```{r}
mtcars$hp
```

Use the `c()` function to make a vector consisting of the course code (`"STAT"` and `545`). Notice the coercion. Vectors must be homogeneous.

```{r}

```


Subset the first entry. Remove the first entry. Note the base-1 system.

```{r}

```

Use `<-` to change the second entry to "545A". Using the same approach, add a third entry, "S01".

```{r}

```

Subset the first and third entry. Order matters! Subset the third and first entry.

```{r}

```

Explore integer sequences, especially negatives and directions. Especially `1:0` that might show up in loops!

```{r}

```

Singletons are also vectors. Check using `is.vector`.

```{r}

```


### Vectorization and Recycling

A key aspect of R is its vectorization. Let's work with the vector following vector:

```{r}
(a <- 7:-2)
(n <- length(a))
```

Square each component:

```{r}

```

Multiply each component by 1 through its length:

```{r}

```

It's important to know that R will silently recycle! Unless the length of one vector is not divisible by the other. Let's see:

```{r}

```


This is true of comparison operators, too. Make a vector of logicals using a comparison operator.

```{r}

```


Now try a boolean operator. Note that && and || are NOT vectorized!

```{r}

```


Recycling works with assignment, too. Replace the entire vector a with 1:2 repeated:

```{r}

```


### Special Subsetting

We can subset vectors by names and logicals, too.

Recall the course vector:

```{r}
course <- c("STAT", "545A", "S01")
```

Let's give the components some names ("subject", "code", and "section") using three methods:

1. Using the setNames function.
- Notice that the vector does not change!!

```{r}

```


2. Using the names function with `<-`. Also, just explore the names function.

```{r}

```

3. Re-constructing the vector, specifying names within `c()`.

```{r}

```


Subset the entry labelled "section" and "subject". 

```{r}

```

Amazingly, we can also subset by a vector of logicals (which will be recycled!). Let's work with our integer sequence vector again:

```{r}
(a <- 7:-2)
(n <- length(a))
```

## Lists

Unlike vectors, which are atomic/homogeneous, a list in R is heterogeneous. 

Try storing the course code (`"STAT"` and `545`) again, but this time in a list. Use the `list()` function.

```{r}

```


Lists can hold pretty much anything, and can also be named. Let's use the following list:

```{r}
(my_list <- list(year=2018, instructor=c("Vincenzo", "Coia"), fav_fun=typeof))
```


Subsetting a list works similarly to vectors. Try subsetting the first element of `my_list`; try subsettig the first _component_ of the list. Notice the difference!

```{r}

```

Try also subsetting by name:

```{r}

```

Smells a little like `data.frame`s. It turns out a `data.frame` is a special type of list:

```{r}
(small_df <- tibble::tibble(x=1:5, y=letters[1:5]))
```

```{r}
is.list(small_df)
```

```{r}
as.list(small_df)
```

Note that there's a difference between a list of one object, and that object itself! This is different from vectors.

```{r}
identical(list(4), 4)
identical(c(4), 4)
```

# Part II

## Global Environment

When you assign variables in R, the variable name and contents are stored in an R environment called a global environment.

See what's in the Global Environment by:

- Executing `ls()`.
- Looking in RStudio, in the "Environments" pane.

```{r}

```


Making an assignment "binds" an object to a name within an environment. For example, writing `a <- 5` assigns the object `5` to the name `a` in the global environment.

The act of "searching for the right object to return" is called scoping.

By the way: the global environment is an object, too! It's the output of `globalenv()`, and is also stored in the variable `.GlobalEnv`:

```{r}
globalenv()
.GlobalEnv
```

## The Search Path

How does R know what `iris` is, yet `iris` does not appear in the global environment? What about functions like `length`, `sum`, and `print` (which are all objects, too)?

Let's explore.

1. Each package has its own environment.
    - Install and load the `pryr` package, and use `ls()` to list its bindings (its name is "package:pryr").

```{r}

```

2. There's a difference between an _environment_ and its _name_. Get the environment with name "package:pryr" using the `as.environment()` function.

```{r}

```


3. Each environment has a parent. Use `parent.env()` to find the parent of the global environment.

```{r}

```


4. There are packages that come pre-loaded with R, and they're loaded in a sequence called the search path. Use `search()` to identify that path; then see it in RStudio.

```{r}

```

First scoping rule: R looks to the parent environment if it can't find an object where it is.

5. Use `pryr::where()` to determine where the first binding to the name `iris` is located.

```{r}

```


6. Override `iris` with, say, a numeric. Now `where()` is it? Can you extract the original?

```{r}

```


7. Override `sum` with, say, a numeric. `where()` is `sum` now? Can you still use the original `sum()` function?

```{r}

```

Special scoping rule for functions! R knows whether or not to look for a function.

8. Look at the source code of the `pryr:where()` function. It contains a line that creates a binding for "env". Why, then, is `env` nowhere to be found? Answer: execution environments. 

```{r}
pryr::where
```



