---
title: "Smoothing time series data using Functional Programming"
output: html_document
---


Enter the `slide()`, `tile()`, and `stretch()` families. Best explained by the [tsibble demo](https://tsibble.tidyverts.org/), specifically [this rolling window animation](https://tsibble.tidyverts.org/reference/figures/animate-1.gif).

```{r}
mauna <- co2 %>% 
	as_tsibble() %>% 
	rename(yearmonth = index,
		   co2 = value) %>% 
	fill_gaps()
ggplot(mauna, aes(yearmonth, co2)) +
	geom_line() +
	theme_bw()
```

To "wash out" the seasonal component (for use in some downstream analysis, such as estimating a trend), we could average-out the data within each year. The `tile()` family makes this easy:

```{r}
(mauna_tiled <- mauna %>% 
 	index_by(year = year(yearmonth)) %>% 
 	summarise(tiled_co2 = tile_dbl(
 		co2, mean,
 		na.rm = TRUE,
 		.size = 12
 	)))
```

But, why just average in chunks defined by January - December? Why not February to the following January? Or March - February? `slide` does all of these.

```{r}
(mauna_washed <- mauna %>% 
	mutate(washed_co2 = slide_dbl(
		co2, mean, 
		na.rm  = TRUE, 
		.size  = 12, 
		.align = "center-left"
	)))
ggplot(mauna_washed, aes(yearmonth)) + 
	geom_line(aes(y = co2), alpha = 0.5) +
	geom_line(aes(y = washed_co2), colour = "blue")
```

