---
title: "cm008 Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## mutate

No need for loops!

`mutate` creates a new variable by calculating from other variables. Let's get GDP by multiplying GPD per capita with population:

```{r}
mutate(gapminder, gdp = gdpPercap * pop)
```

You can define multiple new variables -- even back-dependent on new ones! Let's also create a column for GDP in billions, rounded to one decimal:

```{r}
mutate(gapminder, 
       gdp     = gdpPercap * pop, 
       gdpBill = round(gdp/1000000000, 1))
```

`transmute` works the same way, but drops all other variables. 

__Exercise__: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
mutate(gapminder, cc = paste(country, continent, sep=", "))
```


## `summarize` and `group_by`

`summarize` reduces a tibble according to summary statistics.

```{r}
summarize(gapminder, mean_pop=mean(pop), sd_pop=sd(pop))
```

Not very useful by itself! But, with the `group_by` function, the `summarize` function is very useful:

```{r}
gapminder %>% 
    group_by(country) %>% 
    summarize(mean_pop=mean(pop), sd_pop=sd(pop))
```

The `group_by` function splits the tibble into parts -- in the above case, by country. Notice the "Groups" indicator in the following output:

```{r}
group_by(gapminder, continent, country, year < 1970)
```

Let's get a summary of this grouping:

```{r}
(out1 <- gapminder %>% 
    group_by(continent, country, year < 1970) %>% 
    summarize(mean_pop=mean(pop), sd_pop=sd(pop)))
```

Note that the output is still a tibble, but one "layer" of grouping has been peeled back: the `year < 1970` variable. `summarize` again and you'll see that the tibble was no longer grouped by `year < 1970`. 

```{r}
out1 %>% 
    summarize(mean_pop=mean(mean_pop))
```


__Exercise__: Find the minimum GDP per capita experienced by each country

```{r}
gapminder %>%
    group_by(country) %>% 
    summarize(mingdp=min(gdpPercap))
```


__Exercise__: How many years of record does each country have?

```{r}
gapminder %>%
    group_by(country) %>% 
    summarize(n_distinct(year))
```


__Exercise__: Within Asia, what are the min and max life expectancies experienced in each year?

```{r}
gapminder %>% 
    filter(continent=="Asia") %>% 
    group_by(year) %>% 
    summarize(minexp=min(lifeExp),
              maxexp=max(lifeExp))
```