---
title: 'cm005: `dplyr` Exercise'
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Optional, but recommended startup

For practice, change the file output to both html and md _documents_ (not notebook).

## dplyr

Note that you need to use the assignment operator to make any changes.

### `filter()`

1. Find all entries of Canada and Algeria. 

```{r}
filter(gapminder, 
       country == "Canada" |
           country == "Algeria")
filter(gapminder,
       country %in% c("Canada", "Algeria"))
```


2. Find all entries of Canada and Algeria, occuring in the '60s. 

```{r}
filter(gapminder,
       country %in% c("Canada", "Algeria"), year < 1970, year >= 1960)
```


3. Find all entries of Canada, and entries of Algeria occuring in the '60s. 
```{r}
filter(gapminder,
       (country == "Canada") |
           (country == "Algeria" & 
            year %in% 1960:1969))
```


4. Find all entries _not_ including European countries.

```{r}
filter(gapminder, 
       continent != "Europe")
```

__Exercise__: Take all countries in Europe that have a GPD per capita greater than 10000, and select all variables except `gdpPercap`. (Hint: use `-`).

```{r}
gapminder %>% 
    filter(continent == "Europe",
           gdpPercap > 10000) %>% 
    select(-gdpPercap)
```

Order `gapminder` by population, then life expectancy:

```{r}
arrange(gapminder, pop, lifeExp)
```

__Exercises__:

1. Order the data frame by year, then descending by life expectancy.

```{r}
arrange(gapminder, year, desc(lifeExp))
```


2. In addition to the above exercise, rearrange the variables so that `year` comes first, followed by life expectancy. (Hint: check the documentation for the `select` function for a related handy function).

```{r}
gapminder %>% 
    arrange(year, desc(lifeExp)) %>% 
    select(year, lifeExp, everything())
```

### piping

- take all entries of Canada and Algeria occuring in the '60s, and
- select the `country`, `year`, and `gdpPercap` columns.

We could do...

```{r}
select(filter(gapminder, 
              country %in% c("Canada", "Algeria"), 
              year <= 1969, year >= 1960),
       country, year, gdpPercap)
```

The same operation above becomes:

```{r}
gapminder %>% 
    filter(country %in% c("Canada", "Algeria"), 
           year <= 1969, year >= 1960) %>% 
    select(country, year, gdpPercap)
```
