---
title: "cm007 Exercises: Exploring Aesthetic Mappings"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


# Beyond the x and y aesthetics

For each geometric object, there is a set of aesthetic mappings that must be specified. 

- Heat map, being sure to fill in missing categories.

OLD:

1. Make a scatterplot of `gdpPercap` vs `lifeExp`. Store ~it~ the output of the `ggplot` function in a variable called `p2`.

```{r}
p2 <- ggplot(gapminder, 
             aes(x=gdpPercap,
                 y=lifeExp))
p2 + geom_point()
```


2. To `p2`, make the size of the points indicate the `year`, choose a level of alpha transparency that you're happy with, and make the points your favourite colour.

```{r}
p2 + geom_point(aes(size=year),
                colour="blue", 
                alpha=0.1)
## This *doesn't* work: (why not?)
p2 + geom_point(aes(size=year, colour="blue"), 
                alpha=0.1)
```


3. To `p2`, colour the points by `continent`, ~~but this time~~ with year being represented by the size of the points, like we did in the previous exercise.

```{r}
p2 + geom_point(aes(colour=continent, size=year))
```


4. To `p2`, add another layer called `scale_x_log10()` _in addition to the original `geom_point()` layer_. Make a second plot by redoing the plot in (1), but replacing `gdpPercap` with `log10(gdpPercap)`. What do you notice?

```{r}
p2 + geom_point() + scale_x_log10()
ggplot(gapminder, 
             aes(x=log(gdpPercap),
                 y=lifeExp)) + geom_point()
```