---
title: "cm008 Exercises"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

## `mutate()`

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}

```


Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}

```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}

```


Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}

```

These functions we've seen are called __vectorized functions__. 

## `summarize()` and `group_by()`

Use `summarize()` to compute the mean and median life expectancy using all entries:

```{r}

```

Do the same thing, but try:

1. grouping by country
2. grouping by continent and country

```{r}

```

- Notice the columns that are kept. 
- Notice the grouping listed above the tibble, especially without a call after grouping.
- Notice the peeling of groups for each summarize. 

Question: What if I wanted to keep the other numeric columns (gdpPercap, pop)? Can I? Would this even make sense?

For each continent: What is the smallest country-wide median GDP per capita?

```{r}

```

Note that ggplot2's grouping is different from dplyr's! Try making a spaghetti plot of lifeExp over time for each coutry, by piping in a grouped data frame -- it won't work:

```{r}

```


Your turn! For each continent, what is the median GDP per capita of countries with high (>60) life expectancy vs countries with low (<=60)? Sort this data frame by median GDP per capita.

```{r}

```

There are special functions to summarize by. Let's see some of them:

- `n()`: Number of rows in the group.
- `n_distinct()`

Convenience functions:

- `tally()` (= `summarize(n = n())`)
- `count(...)` (= `group_by(...) %>% tally()`)

n_distinct: How many years of record does each country have?

```{r}

```

count

## Function types

Let's stop to identify some theory of function types, and the `dplyr` approach to them.

| Function type | Explanation | Examples | In `dplyr` |
|------|-----|----|----|
| Vectorized functions | These take a vector, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. | `cos`, `sin`, `log`, `exp`, `round` | `mutate` |
| Aggregate functions | These take a vector, and return a vector of length 1 | `mean`, `sd`, `length`, `typeof` | `summarize`, esp with `group_by`. |
| Window Functions | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag`, `rank`, `cumsum` | `mutate`, esp with `group_by` |

For any generic output, we can use dplyr's `do()` function -- but that's a topic for STAT 547. 

## Grouped `mutate()`

Calculate the growth in population since the first year on record _for each country_. `first()` is useful.

```{r}

```

Notice that `dplyr` has retained the original grouping.

How about growth compared to `1972`?

```{r}

```

Make a new variable `pop_last_time`, as the "lag-1" population -- that is, the population from the previous entry of that country. Use the `lag` function.

```{r}

```

Similar: `lead` function.

Notice the NA's. 

## Putting it all together

Your turn: Use what we learned to answer the following questions.

1. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent.

```{r}

```

2. Compute the relative gdp (NOT per capita!) of each country compared to Canada (= GDP of a country / GDP of Canada).

```{r}

```

Sanity check: are Canada's numbers = 1? What is the spread of numbers like (should be small)?

## Summary of major one-table functions

- `select()`
- `filter()`
- `arrange()`
- `mutate()`
- `summarize()`

Together with `group_by()` and "twists" of the above. 


## Practice Exercises

Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.

1. Convert the population to a number in billions.

2. Compute the change in population from 1962 to 1972 for each country.

3. Rank the continents by GDP per capita. You should have two columns: one with continent, and another with the ranking (1 through 5). __Hint__: use the `rank()` or `min_rank()` function.
