---
title: "Class Meeting 11: Getting Data from the Web, Part 1"
output:
    html_notebook:
        toc: true
        theme: cerulean
        number_sections: true
editor_options: 
  chunk_output_type: inline
---

# Motivation

Today's topic: scraping the web.

This means reading webpages like [this wikipedia page](https://en.wikipedia.org/wiki/List_of_roller_coaster_rankings) and putting relevant info into R.

# Getting Started

1. Install [`rvest`](https://github.com/hadley/rvest): `install.packages("rvest")`.

2. Load it:

```{r}
library(tidyverse)
library(rvest)
```

3. Download [cm111-simple_script.html](cm111-simple_script.html) to your participation repo.

4. Explore the html script, and the structure of html.


# Scraping with `rvest`: Overview

Overall workflow with `rvest`:

1. Read in a webpage with `read_html()`.
2. Select parts of the page with `html_nodes()`.
	- Optional, especially when reading tables.
3. Convert parts of the selected html to R objects:
	- text: `html_text()` for the actual text between the tags, but also relevant are:
		- `html_name()` for the name of the (highest level) tag itself,
		- `html_attrs()`/`html_attr()` for the attributes.
	- tibbles: `html_table()`

## Together:

We'll use `rvest` to read in parts of `cm111-simple_script.html` into R.

1. Apply `read_html()` to the file: reads in the page, outputs a special R object.

```{r}
simple_page <- FILL_THIS_IN
```

2. Pipe the output into `html_structure()`: will print the html to screen.

```{r}
simple_page %>% html_structure()
```

3. Apply `html_text()` to `simple_page`: gives me all the text (not the tags).

```{r}

```

4. `html_nodes()`: selects only relevant parts of the html, constructing a (special) list of all relevant pieces. 
	- Example: select all text between paragraph tags.
	- Example: select all hyperlink text.
	- Example: select all tags with the "wiki" id (Hint: `#wiki`)
5. Then, pipe into `html_text()`.

```{r}
simple_page %>% 
	html_nodes(FILL_THIS_IN)
simple_page %>% 
	html_nodes(FILL_THIS_IN)
simple_page %>% 
	html_nodes(FILL_THIS_IN)
```


## Your Turn, Round 1

Practice with CSS selection! Partner up, and see how many levels you can complete in the [CSS selector game](http://flukeout.github.io/).

## Your Turn, Round 2

Return the text in `simple_page` contained within:

1. all `<a>` tags within `<p>` tags.

```{r}
simple_page %>% 
	html_nodes(FILL_THIS_IN) %>% 
	FILL_THIS_IN
```

2. any tag.

```{r}
simple_page %>% 
	html_nodes(FILL_THIS_IN) %>% 
	FILL_THIS_IN
```


## Reading the tags

We can extract other things besides the text between the tags:

- `html_name()` gives the name of the tags. Try it with `html_nodes(*)`.

```{r}
simple_page %>% 
	html_nodes("*")
```

- `html_attrs()` gives all attributes of the tags. Try it with `html_nodes("a")`:

```{r}
simple_page %>% 
	html_nodes("a")
```

- `html_attr()` (not plural) extracts a vector of the attribute contents by the `name` argument. Try `name="href"` with `html_nodes("a")`:

```{r}
simple_page %>% 
	html_nodes("a")
```


- Make a tibble of hyperlink text and their associated url's:

```{r}
tibble(
	text = FILL_THIS_IN,
	url  = FILL_THIS_IN
)
```

## When CSS selection gets harder

Let's extract the songs that appear on [Kane Brown's page on musixmatch](https://www.musixmatch.com/search/kane%20brown).

Viewing the source to find the CSS is horrendous. In comes [`SelectorGadget`](https://selectorgadget.com/)! Visit the page to install the gadget.

1. Use the gadget to select the songs. 
2. Put the CSS selection in the `html_nodes()`.

```{r}

```

## Reading a table

Goal: extract the Wikipedia table of the [tallest steel roller coasters](https://en.wikipedia.org/wiki/List_of_roller_coaster_rankings#Tallest_steel_roller_coasters) in the world with `html_table()`.

```{r}
read_html("https://en.wikipedia.org/wiki/List_of_roller_coaster_rankings") %>% 
	FILL_THIS_IN
```


# Data structures and Importing

We've already seen `readr`, but what about importing other data types?

- JSON, with the `jsonlite` package. Examples from the `fromJSON()` documentation:

```{r}
fromJSON('{"city" : "Z\\u00FCrich"}') # A list
fromJSON("https://api.github.com/users/hadley/orgs") %>% 
	glimpse() # A data frame
toJSON(mtcars, pretty=TRUE)
```


- XML, with the `xml2` package (similar to reading webpages).
	- [Example from w3cshools](https://www.w3schools.com/xml/simple.xml)

```{r}
read_xml("https://www.w3schools.com/xml/simple.xml") %>% 
	as_list()
```


Others:

- [`googledrive`](https://googledrive.tidyverse.org/)
- See the [tidyverse listing of import packages](https://www.tidyverse.org/packages/#import).