---
title: "Class Meeting 112: Getting Data from the Web, Part 2: API's"
output:
    html_notebook:
        toc: true
        theme: cerulean
        number_sections: true
editor_options: 
  chunk_output_type: inline
---

## Announcements

- Homework 10 is now optional. Also, one pair of peer reviews.
- My office hour this week is after class today.

## Objective

Objectives of today's lesson:

- Form a web API request to retrieve data
- Make an API request from within R, esp. with the `httr`.
- Familiarize yourself with the context of forming these requests: API's in general, and HTTP.

## Lesson Overview

Starting from the bottom objective, let's set the context:

1. What is an API? Plus requests.
2. What is HTTP? Requests and Responses.

Then, let's focus on the data retrieval task:

3. Using the `httr` package to form requests, and specialized packages.
4. Other relevant examples of API's

We'll wrap up with: 

5. Instructor/course evaluations

## Participation

Download the `.Rmd` version of this very document.

## Overall Resources

- The [2016 ROpenSci tutorial](https://github.com/ropensci-training/user2016-tutorial).
	- Parts 1 and 2 provide a good overview of today's topic.
	- Part 3 covers scraping (we covered that last class).
- ["An Introduction to APIs"](https://zapier.com/learn/apis/), a brief book by Brian Cooksey. 	
	- Concise yet complete look at today's concepts.

I'll post more specific resources at the start of each topic.

# What is an API?

API = Application Programming Interface

API's deliver information from a server by request.

## Resources

I suggest these resources to get oriented:

- ["What is an API? In English, please."](https://medium.freecodecamp.org/what-is-an-api-in-english-please-b880a3214a82), by Petr Gazarov
	- Fantastic explanation and orientation of what API's are.
- [What is an API?](https://www.youtube.com/watch?v=s7wmiS2mSXY), by MuleSoft Videos
	- Brief, high-level YouTube video
- The [ROpenSci 2016 tutorial](https://github.com/ropensci-training/user2016-tutorial/blob/master/slides.pdf), Part 1, linked near the top of the page. 
	- Nice walk-through. You might find it useful to first get a higher-level overview from the above two resources.

## Demonstrations of API's

__Visiting a website__

1. You enter a url (the _request_), say <https://twitter.com/>, into your web browser (the _client_).
2. The Twitter server delivers the requested page in the form of html (the _response_).
	- The _transfer_ is via `https`: `h`yper`t`ext `t`ransfer `p`rotocol (`s`ecure).
3. Your browser displays the html.

__Google Flights__

1. Pick a flight from https://www.google.com/flights
2. Google sends your browser information about relevant flights.

Notice:

- Notice the _query_ at the end of the url!
- Many API's, like google flights, use _third-party API's_ -- it uses WestJet, AirCanada, etc.

__GitHub API__

Can request "data"; usually comes in json or xml, not html!

1. Request, say, information about the GitHub Orgs that Hadley belongs to:
	- `https://api.github.com/users/hadley/orgs`
2. The GitHub API returns not html, but JSON.

__OMDb__

API's are not always open, and require an _API key_.

Let's request info about Jurassic Park from OMDb. We'll also start writing requests with this one. (You'll need to register for an API key if you want to follow)

1. Go to http://www.omdbapi.com/
2. Request info about Jurassic Park
	- The syntax is specific to OMDb
	
__OMDb Exercise__:

1. Register for an API key at http://www.omdbapi.com/
2. Form a request!

```{r}
s
```


__Google API's__

Some services offer _many_ API's. 

1. Check out the API's available [from Google](https://developers.google.com/).
	- Example: [Google Maps Directions API](https://developers.google.com/maps/documentation/directions/start)
	- `maps.googleapis.com`, "the other side" of `maps.google.ca`.

__Local API's__

API's are not always on another server. Local examples:

- Your operating system
- An R package
- Shiny apps (not local if you release them to the web).

# What is HTTP?

HTTP = hypertext transfer protocol. A standardized format for communicating between servers.

## Resources

I suggest ["What is HTTP?"](https://www.webnots.com/what-is-http/) by webnots. Gets right to it, in an uncomplicated way.

And, for info on HTTP verbs/methods, the [REST API tutorial](https://www.restapitutorial.com/lessons/httpmethods.html)

## Getting Oriented

What really happens when you type a url into your browser?

1. You type a url in the address bar of your browser.
2. An http _request_ is formed, and sent to the website's server.
	- Components are formed using defaults and using components of the url.
3. The server sends an http _response_.

The _request_ and _response_ are plain text, and each have a specific structure and components.

## Revealing the request and response

Let's view uncover this "behind the scenes" messaging that's going on. 

- __Using the terminal__: Use `curl -v`. 
	- Example `curl -v https://github.com/STAT545-UBC/Classroom`. The request starts with `>`, and the response starts with `<`.
- __Using your Browser__: You can at least do this with Google Chrome. See the [webnots article](https://www.webnots.com/what-is-http/) I link to above.

## Request and response: components

Here are some noteworthy components to look out for:

__Response__ 

- _Status_: "404" and others 
	- See this [Wikipedia description](https://en.wikipedia.org/wiki/List_of_HTTP_status_codes) of status codes; also described in the [webnots article](https://www.webnots.com/what-is-http/) referenced earlier.
- _Headers_: info about the page itself.
- _Content_: The actual content of the page. 

__Request__

- _url_
- _method_, aka _HTTP verb_ (more on this in the next section)
	- Default verb: `GET`.
- _headers_ allow you to modify the request.

## HTTP verbs/methods

By default, your browser uses `GET`. You can interact (and actually _modify_) the server using other verbs. Common ones are

- `GET` -- retrieve info from the server.
	- Most useful for us -- gets data!
- `POST` -- create content on the server.
- `PUT` -- update content on the server.
- `DELETE` -- delete content from the server.

# Using R to make requests

The [`httr` R package](https://github.com/r-lib/httr) provides an interface for HTTP verbs. We'll primarily look at this.

```{r}
library(httr)
```

__Acronym check-in__: Note that these acronyms look similar, but are not the same: html != http != httr.

## Resources

Suggested resources (which also briefly explain what API's are):

- Tyler Clavelle's [Using R to extract data from web APIs](https://tclavelle.github.io/blog/r_and_apis/)
- The [ROpenSci 2016 tutorial](https://github.com/ropensci-training/user2016-tutorial/blob/master/slides.pdf), Part 2, linked near the top of the page. 
- The [`httr` quickstart vignette](https://cran.r-project.org/web/packages/httr/vignettes/quickstart.html)
	- Quick and dirty "how to use" guide. Good if you're comfortable enough with API's and HTTP, and want to adapt to the `httr` package.


## HTTP verbs/methods: Through `httr`

Those same HTTP verbs are `httr` functions. We'll be focussing on `GET()`!

__Together__:

1. `GET()` a website, any website.
2. Inspect its `str()`ucture, `class()`, and `names()`.

```{r}
site <- GET("PUT_URL_HERE")
```

3. What's the `$status_code`? How about the `$content`?

```{r}

```

4. Try the function versions: `status_code()` and `content()`. 
	- Is the output of `content()` html? You can scrape with `rvest`.
	- Try again from the top, this time with Hadley's GitHub Orgs: `https://api.github.com/users/hadley/orgs`

```{r}
status_code(site)
content(site)
```

5. The GitHub API (Hadley's GitHub Orgs) returns JSON. Why not just use `jsonlite::fromJSON()`?
	- Sometimes data come in XML format, and can't do this.

```{r}
jsonlite::fromJSON("https://api.github.com/users/hadley/orgs")
```

## Exercise

1. Collect the information about a movie of your choice from [the OMDb](http://www.omdbapi.com/). You'll have to sign up for an API key (it's free).
2. Extract the `Plot`.

```{r}

```


## API wrappers and examples

Many R packages are wrappers to an API. Examples:

- `ggmap` accesses [Google Maps API](https://developers.google.com/maps/documentation/).
- `rebird` accesses the [ebird](https://ebird.org/home) database.
- `geonames` accesses the [geonames.org](geonames.org) API.
- `plotly` accesses the [plotly API](https://plot.ly/graphing-libraries/).
- `rplos` accesses the [Public Library of Science](https://www.plos.org/) API.

Andrew MacDonald wrote an overview of a few of these in the [2015 STAT 547 cm111 notes, "Install and Play" section](http://stat545.com/webdata02_activity.html#install-and-play).

In addition, here are other API's to be aware of (some of which have an R package, I believe):

- [Facebook API](https://developers.facebook.com/)
- [Twitter API](https://developer.twitter.com/)
- [Star Wars API](https://swapi.co)
- [API of Ice and Fire](https://anapioficeandfire.com)
- [OpenWeatherMap](https://openweathermap.org/api)
- [Open Movie Database](https://www.omdbapi.com)
- [Quandl API](https://www.quandl.com/)

And a huge list of API's can be found [here](http://www.programmableweb.com/category/all/apis).

# Course/Instructor Evaluations

Please take 5 minutes to fill out the course/instructor evaluations!